\myfont Implementation and evaluation of register allocation for translating LLVM-{}- to x86-64

### 2024-01-13 article.tex
Mostly coalescing code

Because of this, for any phi node of the form 
\begin{equation}
i a = \Phi (b_1, ..., b_n) \label{phi}
\end{equation}
add edges to all \textit{live-out} variables not in \(\{b_1, ..., b_n\}\)
\[
  \forall b_j \in \mathit{out}[i] \setminus \{b_1, ..., b_n\}, \text{add\_edge}(a, b_j),
\]
and for any other instruction that defines a variable \(a\)
\[
  \forall b_j \in \{b_1, ..., b_n\}, \text{add\_edge}(a, b_j).
\]
Although the Appel text \cite{tiger} describes interference of variables with concrete registers as well as overlapping variables, this isn't considered in this implementation % FIXME: why??
for simplicity's sake.

@@

% Side by side figures 
\begin{figure}[H]
  \centering
  \begin{minipage}[b]{0.31\linewidth}
    \begin{subfigure}[b]{0.70\textwidth}
      \includesvg[height=9em]{countdown.simple2}
      \caption{Graph}\label{fig:countdown.dot2}
    \end{subfigure}
    \begin{subfigure}[b]{0.27\textwidth}
      \centering
        \texttt{\%c1}
        \vspace{0.5em}
      \caption{Stack}\label{fig:countdown.stack2}
    \end{subfigure}
    \caption{Push \texttt{\%c1}}
  \end{minipage}%
  \begin{minipage}[b]{0.31\linewidth}
    \begin{subfigure}[b]{0.70\textwidth}
      \includesvg[height=9em]{countdown.simple3}
      \caption{Graph}\label{fig:countdown.dot3}
    \end{subfigure}
    \begin{subfigure}[b]{0.27\textwidth}
      \centering
        \texttt{\%argv}\\
        \texttt{\%c1}
        \vspace{0.5em}
      \caption{Stack}\label{fig:countdown.stack3}
    \end{subfigure}
    \caption{Push \texttt{\%argv}}
  \end{minipage}%
  \begin{minipage}[b]{0.31\linewidth}
    \begin{subfigure}[b]{0.70\textwidth}
      \includesvg[height=5.5em]{countdown.simple4}
      \caption{Graph}\label{fig:countdown.dot4}
    \end{subfigure}
    \begin{subfigure}[b]{0.27\textwidth}
      \caption{Stack}\label{fig:countdown.stack4}
    \end{subfigure}
    \caption{Spill \texttt{\%c2}}
  \end{minipage}%
\end{figure}
\noindent From this point onwards  nothing can be done to further simplify the graph as all of the remaining nodes are of significant degree (\(\geq 2\)). Therefore, one of the remaining nodes is spilled to memory and entirely removed from the graph as can be seen on Figure \ref{fig:countdown.dot4} and the stack reset as can be seen on Figure \ref{fig:countdown.stack4}.

@@

\noindent The purpose is only to assign the same registers to different variables, which stands to benefit from less \texttt{mov} instructions being used.
%. You don't need to note down information about blocks etc.

@@

@@

The instruction set used in this paper will be a union of the sets used in the 2022 and 2023 compilers courses in order to work as a drop-in replacement of LLVM for either of the two respective source languages:  Tiger and Dolphin. This instruction set is a subset of the one used in practice, as, for instance, neither of the languages implemented support exception handling, floating point operations and so on, and instead only strive to cover the basics of compilers.

@@

The testing framework is roughly based on the build system that came with the Tiger work environment. It is based on a collection of files that can be compiled and executed on their own. Each of these are enumerated in the test runner (\texttt{bin/testrunner.ml}) with parameters associated.   Each of these are then compiled with the \texttt{clang} compiler and executed, capturing all the relevant output data, which is stdout and exit code. After this, the same test file is compiled for each of the allocators. The executables generated are then also executed, also storing the output data.   All test cases are deterministic and will halt at some point. This is necessary because tests can easily end in an infinite loop because of erroneous translation. As each of these halt or time out, their output is compared to what was output from executables generated by the 'correct' compiler of \texttt{clang}.

>>>>

Because of this, for any phi node of the form 
\begin{equation}
 a = \Phi (b_1, ..., b_n) \label{phi}
\end{equation}
add edges to all \textit{live-out} variables not in \(\{b_1, ..., b_n\}\)
\[
  \forall b_j \in \mathit{out}[i] \setminus \{b_1, ..., b_n\}, \text{add\_edge}(a, b_j),
\]
and for any other instruction that defines a variable \(a\)
\[
  \forall b_j \in \{b_1, ..., b_n\}, \text{add\_edge}(a, b_j).
\]
Although the Appel text \cite{tiger} describes interference of variables with concrete registers as well as overlapping variables, this isn't considered in this implementation % FIXME: why??
for simplicity's sake.

@@

% Side by side figures 
\begin{figure}[H]
  \centering
  \begin{minipage}[b]{0.31\linewidth}
    \begin{subfigure}[b]{0.70\textwidth}
      \includesvg[height=9em]{countdown.simple2}
      \caption{Graph}\label{fig:countdown.dot2}
    \end{subfigure}
    \begin{subfigure}[b]{0.27\textwidth}
      \centering
        \texttt{\%c1}
        \vspace{0.5em}
      \caption{Stack}\label{fig:countdown.stack2}
    \end{subfigure}
    \caption{Push \texttt{\%c1}}
  \end{minipage}%
  \begin{minipage}[b]{0.31\linewidth}
    \begin{subfigure}[b]{0.70\textwidth}
      \includesvg[height=9em]{countdown.simple3}
      \caption{Graph}\label{fig:countdown.dot3}
    \end{subfigure}
    \begin{subfigure}[b]{0.27\textwidth}
      \centering
        \texttt{\%argv}\\
        \texttt{\%c1}
        \vspace{0.5em}
      \caption{Stack}\label{fig:countdown.stack3}
    \end{subfigure}
    \caption{Push \texttt{\%argv}}
  \end{minipage}%
  \begin{minipage}[b]{0.31\linewidth}
    \begin{subfigure}[b]{0.70\textwidth}
      \includesvg[height=5.5em]{countdown.simple4}
      \caption{Graph}\label{fig:countdown.dot4}
    \end{subfigure}
    \begin{subfigure}[b]{0.27\textwidth}
      \caption{Stack}\label{fig:countdown.stack4}
    \end{subfigure}
    \caption{Spill \texttt{\%c2}}
  \end{minipage}%
\end{figure}
\noindent From this point on  nothing can be done to further simplify the graph as all of the remaining nodes are of significant degree (\(\geq 2\)). Therefore, one of the remaining nodes is spilled to memory and entirely removed from the graph as can be seen on Figure \ref{fig:countdown.dot4} and the stack reset as can be seen on Figure \ref{fig:countdown.stack4}.

@@

\noindent The purpose is only to assign the same registers to different variables, which stands to benefit from less \texttt{mov} instructions being used.
\subsection{Implementation}

The first step is deriving all preference edges:

\begin{figure}[H]
  \centering
  \begin{minted}[linenos,bgcolor=LightGray]{ocaml}
let prefer (insns : Cfg.insn list) : S.SS.t S.ST.t =
  let insn t = function
    | Cfg.Insn (Some d, Ll.PhiNode (_, ops)) ->
        List.fold_left
          (fun t o ->
            match o with
            | Ll.Id sop, _ ->
                S.ST.update sop
                  (function
                    | Some s -> Some (S.SS.add d s)
                    | None -> Some (S.SS.singleton d))
                  t
            | _ -> t)
          t ops
    | _ -> t
  in
  List.fold_left insn S.ST.empty insns
  \end{minted}
  \caption{Implementation of \texttt{prefer} in \texttt{lib/coalesce.ml}}\label{fig:prefer.ml}
\end{figure}
\noindent This is done simply by iterating over all of the instructions, and for every phi node, add a preference edge between \(a\) on the left-hand side of \eqref{phi} and all \(b_i\) on the right. Of note is the fact that only operands of the form \(\%i\) (i.e. \texttt{Id} variants of the \texttt{operand} type) are considered. This is because immediate values may also be specified in an LLVM phi node, whose \texttt{mov} instruction  don't make sense to eliminate. % FIXME: is this true?

Briggs coalescing is then implemented as follows:
\begin{figure}[H]
  \centering
  \begin{minted}[linenos,bgcolor=LightGray]{ocaml}
let coalesce_briggs k (prefs : S.SS.t S.ST.t)
    ((l, g) : Lva.G.V.t S.ST.t * Lva.G.t) : Lva.G.V.t S.table * Lva.G.t =
  let try_coalesce sop dop (l, g) =
    if not (Lva.G.mem_edge g (S.ST.find sop l) (S.ST.find dop l)) then
      let sneighs = S.ST.find sop l |> Lva.G.succ g |> Lva.VS.of_list
      and dneighs = S.ST.find dop l |> Lva.G.succ g |> Lva.VS.of_list in
      let neighs = Lva.VS.union sneighs dneighs in
      let sign v = Lva.G.succ g v |> List.length >= k in
      let signeighs = Lva.VS.filter sign neighs in
      if Lva.VS.cardinal signeighs < k then
        coalesce (S.ST.find sop l) (S.ST.find dop l) (l, g)
      else (l, g)
    else (l, g)
  in
  let try_pref sop dops (l, g) = S.SS.fold (try_coalesce sop) dops (l, g) in
  S.ST.fold try_pref prefs (l, g)
  \end{minted}
  \caption{Implementation of \texttt{coalesce\_briggs} in \texttt{lib/coalesce.ml}}\label{fig:coalesce_briggs.ml}
\end{figure}
\noindent For every preference edge, attempt to coalesce them into one. The \texttt{try\_coalesce} function is applied to every pair \((a,b)\) for all \(\Phi\text{-nodes}\).
At first whether \(a\) and \(b\) are in conflict is checked, because only nodes that are not in interference may be coalesced, and if they are there is no point in finding their neighbors.
The degree of potential node \(ab\) is taking all of the neighbors  of \(a\) and \(b\), and if the union of these is of magnitude \(\geq k\) it doesn't satisfy the Briggs criteria and no further action is taken. If it is \(<k\), i.e. node \(ab\) is of insignificant degree, the two are combined in the \texttt{coalesce} function.

%. You don't need to note down information about blocks etc.

@@

The following is a very tedious implementation, stemming from the the \texttt{ocamlgraph} library simply doesn't support contracting a graph. Instead one is directed towards constructing a new graph and copying over the vertices and edges in that order. This is done by first folding over all existing vertices,  adding the one's that are not in the process of being coalesced, and when the first of the pair being coalesced is met, set the third tuple entry to be some, the associated data being the vertex identifier as well as it's associated symbols. Each vertex maps to a symbol set (i.e. \texttt{S.SS.t}) of variables already coalesced which is initially a singleton.

\begin{figure}[H]
  \centering
  \begin{minted}[linenos,bgcolor=LightGray]{ocaml}
let coalesce v1 v2 (st, g) =
  if v1 = v2 then (st, g)
  else
    let g' = Lva.G.create () in
    let st', vt', _ =
      Lva.G.fold_vertex
        (fun v (st, vt, jn) ->
          if v = v1 || v = v2 then (
            match jn with
            | None ->
                let s = Lva.G.V.label v in
                (st, vt, Some (v, s))
            | Some (v1, s1) ->
                let s' = Lva.G.V.label v |> S.SS.union s1 in
                let v' = Lva.G.V.create s' in
                Lva.G.add_vertex g' v';
                ( S.SS.fold (fun e l -> S.ST.add e v' l) s' st,
                  Lva.VT.add v1 v' vt |> Lva.VT.add v v',
                  None ))
          else
            let s = Lva.G.V.label v in
            let v' = Lva.G.V.create s in
            Lva.G.add_vertex g' v';
            ( S.SS.fold (fun e st -> S.ST.add e v' st) s st,
              Lva.VT.add v v' vt,
              jn ))
        g
        (S.ST.empty, Lva.VT.empty, None)
    in
    Lva.G.iter_edges
      (fun v1 v2 ->
        let v1 = Lva.VT.find v1 vt' in
        let v2 = Lva.VT.find v2 vt' in
        if v1 <> v2 then Lva.G.add_edge g' v1 v2)
      g;
    (st', g')
  \end{minted}
  \caption{Implementation of \texttt{coalesce} in \texttt{lib/coalesce.ml}}\label{fig:coalesce.ml}
\end{figure}

@@

The instruction set used in this paper will be a union of the sets used in the 2022 and 2023 compilers courses in order to work as a drop-in replacement of LLVM for either of the two respective source languages:  Tiger and Dolphin. This instruction set is a subset of the one used in practice, as, for instance, neither of the languages implemented support interrupts, floating point operations and so on, and instead only strive to cover the basics of compilers.

@@

The testing framework is roughly based on the build system that came with the Tiger development environment. It is based on a collection of files that can be compiled and executed on their own. Each of these are enumerated in the test runner (\texttt{bin/testrunner.ml}) with parameters associated.   Each of these are then compiled with the \texttt{clang} compiler and executed, capturing all the relevant output, which is stdout and exit code. After this, the same test file is compiled for each of the allocators. The executables generated are then also executed, also storing the output data.   All test cases are deterministic and will be  halted by configurable timeout at some point. This is necessary because tests can easily end in an infinite loop because of erroneous translation. As each of these halt or time out, their output is compared to what was output from executables generated by the 'correct' compiler of \texttt{clang}.
